---
title: "data_visualization"
author: "Keli Fine"
date: "3/19/2019"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Sources:  
* http://daviddalpiaz.github.io/appliedstats/model-diagnostics.html  
* https://www.oreilly.com/library/view/practical-statistics-for/9781491952955/ch01.html
* https://r4ds.had.co.nz/data-visualisation.html

Libraries:  
* library(corrplot)
* library(descr)
* library(ggplot)
* library(tidyverse)  

### QQ plots

* qqnorm(): quantile-quantile plot for normality
* qqline(): adds line to qqnorm

```{r cars}
data("mtcars")
lm = lm(mtcars$mpg ~ mtcars$cyl, data=mtcars)
qqnorm(resid(lm))
qqline(resid(lm))
```

### Barplots, boxplots, and histograms

* boxplot(variable-variable, data, ylab=''): returns boxplot of specified variable
* hist(df[[variable]], breaks=breaks): returns histogram. freq=false for density line
* lines(density(df[[variable]])): adds density line to plot
* barplot(as.matrix(df))

```{r}
hist(mtcars$mpg, breaks=12)
```
### Frequency, correlation, and contingency

1. breaks <- seq(from=min(df[[variable]]), to=max(df[[variable]]), length=)
2. df <- cut(df[[variable]], breaks=breaks, right=True, include.lowest=True)
3. table(df): returns frequency table
* corrplot(cor(df), method): returns correlation plot
* plot(x, y, xlab, ylab): returns scatterplot
* CrossTable(variable, variable, prop.c=False, prop.chisq=False, prop.t=False): returns contingency table

```{r}
library(corrplot)
corrplot(cor(mtcars))
```
### GGPlot library

* ggplot(data, mapping=aes(x,y, color, size, shape, alpha, fill), position): aes maps by variable, outside by name
* geom_point(): scatterplot
* geom_smooth(): line plot
* geom_bar(): bar plot
* geom_boxplot(): box plot
* geom_polygon(): returns shape plot
* geom_abline(): adds line
* coord_flip(): switches x and y axes
* coord_polar(): uses polar coordinates
* coord_quickmap(): sets aspect ratio for maps
* stat_<func>(): interchangeable with geoms
* facet_wrap(~variable, nrow): returns rows of plots
* facet_grid(variable~variable): returns grid of plots

```{r}
library(tidyverse)
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

