---
title: "data_wrangling"
author: "Keli Fine"
date: "3/19/2019"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(dplyr)
library(nycflights13)
```
Sources:  
* Class lectures week 4 and 5
* https://bookdown.org/rdpeng/exdata/managing-data-frames-with-the-dplyr-package.html


###Regular Expressions

* str_detect(string, pattern): returns True if pattern is found
* str_extract(string, pattern): returns character vector of first match
* str_extract_all(string, pattern): returns character vector of all matches
* str_replace(string, pattern, replacement): replaces first instance of match
* str_replace_all(string, pattern, replacement): replaces all instances of match
* str_split(pattern): splits string by pattern


```{r regex}
sentence <- "The fat cat in the hat sat on the mat."
str_extract_all(sentence, '.at')
str_replace_all(sentence, 'cat', 'dog')
```
###Data transformation

####Tidyr:
* gather(): transforms wide table into long table
* separate(): teases variables apart using splitting patterns
* spread(): transforms long table into wide table

###Dplyr:
* %>%: piping operator
* select(df, column:column): returns subset of columns, can also use starts_with or * ends_with to select columns
* filter(df, variable <>= val): returns subset of rows
* arrange(df, variable): orders results by variable, wrap with desc() for reverse order
* rename(df, variable=newname): renames column of dataframe
* mutate(df, name=func(variable)): adds new column to df
* transmute(df, name=func(variable)): same as mutate but drops all non transformed variables
* group_by(df, column): groups rows by specified variable
* summarize(df, name=func(variable)): returns specified summary statistics 





```{r nycflights}
flights %>% filter(month==1) %>% arrange(day) %>% select(month:arr_delay) %>% mutate(gain=arr_delay-dep_delay)
```
